% Acknowledgements ============================================================

First and foremost I’d like to thank my supervisor Christoph, who taught me how to do research and how to navigate the hectic world of academia. When I started working with you I did not even know how to run a Python script, let alone training a neural network. And here I am, four years later, with a thesis on machine learning for astrophysical data analysis. Thank you for your trust, patient guidance, and complete support. I’m especially grateful for the freedom you gave me to pursue and explore various projects, often inspired by the many ideas you generate like a volcano.

Being at GRAPPA over the last four years has been an amazing experience. I'd like to thank all GRAPPA faculty --- Gianfranco, Samaya, Philipp, Shin'ichiro, Jacco, Ben, and Sera --- for fostering a thriving workplace with your availability, openness, and enthusiasm. A special thank goes to my independent supervisor, Wouter Waalewijn, who has always ensured that I felt comfortable and free to talk. No institute can succeed without the tireless effort of the administration staff. In particular, Ji\v{r}ina \v{S}\'{a}lkov\'{a}, GRAPPA is lucky to have you and I’m hugely grateful for your availability and help in navigating me through life at a scientific institute.

All the past and present GRAPPA PhD students deserve a special thank you for being such wonderful office mates: Ariane, Banafsheh, Ben, Uddipta, Gimmy, Dion, Fabian, Mathis, Oleg, Youyou, Alexandra, Kate, Huyfang, and more recently Pravita and Joy. 
Gimmy, abbiamo iniziato il dottorato lo stesso giorno, lo finiamo insieme. Mi mancheranno le nostre passeggiate senza meta (ma con tappa gelato) e i giri in bici in cui più di una volta ho confuso destra e sinistra\dots\ ma non il tuo ``risotto" agli spinaci! Non vedo l'ora di organizzare le prossime camminate, possibilmente in salita.
From the moment I first entered the GRAPPA PhD office, Uddipta welcomed me with his characteristic warmth. Your kindness never fades — thank you for all the great times we've shared and your adventurous stories.
Fabian, thank you for making our lunch breaks so much more engaging with insightful discussions. I'll greatly miss our chats.
Special thanks goes to my paranymph Dion for translating my English Summary to the Dutch Samenvatting, and for the fun times we shared at the GGI school. 

\clearpage
To the current and past postdocs at GRAPPA --- Sam, Pippa, Camila, Oscar, Pablo, James, Guillermo, Malte --- thank you for all your advice, feedback (especially during application season!), and the postdoc-wisdom you have shared. And especially thank you for all the fun memories and the pub-quizzes! In particular, Camila, you have been a wonderful daily supervisor. Thank you for guiding me through the analysis of cosmological simulations.
James, since you arrived, you have been a constant source of guidance for me. You taught me a lot about science and what it means to be a scientist through our numerous discussions, for which I’m very grateful. Your insightful ability to communicate - and willingness to argue - has been eye-opening. Thank you for your time and dedication to our research; I look forward to what the future holds.

To my first collaborators, Adam and Kosio, thank you a lot for introducing me to the magical world of lensing and helping me in my first steps in learning how to make a machine learn. I have learnt a great deal from both of you.

During my PhD, I had the pleasure of meeting and collaborating with many other great people. Special thanks to Florian, for introducing me to cosmology and for a fun coding week in Annecy, and to Christopher, for introducing me to the gamma-ray sky analysis and its wonders. Another big thanks goes to Francesca, who was a wonderful host at LAPTh in Annecy.

Outside of academia, there are many people that have made my life full, and contributed just as much through their presence.

Fra, dai giorni in cui vivevamo nel mio studio senza mobili, a quelli sul tuo terrazzo con vista giraffe, alle ultime passeggiate lungo i canali, sei stata il mio pezzo di casa fuori casa. Sono molto felice di aver potuto condividere la maggior parte di questi anni ad Amsterdam con te. 
Alle nostre `Orchidee' --- Cha, Eri, Sof, Agne ed Ely --- grazie per esserci venute a trovare portando con voi altri pezzi di casa. 

Marcelo and Celina, you were the first ones to welcome me in Amsterdam, in Esnoga, and in your house. Thank you for always hosting me during many Hagim. 
Emmanuel, thank you for our Amsterdam strolls and inspiring conversations. Though we don’t see each other often, I value how close we’ve stayed since arriving in Amsterdam around the same time.

La `LaMamma', credo che l'articolo --- parte integrante di come mi rivolgo a te --- dica molte cose. Aggiungo un grazie per avermi insegnato l'importanza dell'attenzione e della cura, in ogni loro aspetto e manifestazione. Papà, se ho difficoltà a compiere scelte è merito tuo, che mi hai mostrato e sempre offerto quante possibilità esistono e la libertà di coglierle. Gio, grazie di aver badato a casa e specialmente alla Sbisu. 

\clearpage
Una presenza costante nonostante la distanza --- Gio, non vedo l'ora di continuare a camminare insieme. 

\vskip 2pt

\begin{flushright}
\textit{Noemi, \\
October 2024}
\end{flushright}

