% Publication list ============================================================

{\scshape This thesis is based on the following publications:} \vskip 20pt


\begin{itemize}[]

{\fontfamily{cmr}\selectfont

\item [\cite{Montel:2022fhv}]

{N. Anau Montel}, A. Coogan, C. Correa, K. Karchev, C. Weniger, \textit{Estimating the warm dark matter mass from strong lensing images with truncated marginal neural ratio estimation}, 
\href{https://doi.org/10.1093/mnras/stad2925}{Mon.Not.Roy.Astron.Soc. 527 (2024) 66}, \href{https://arxiv.org/abs/2209.09918}{\ttfamily arXiv:2209.09918}. \vskip 5pt

{All authors participated in the planning of the project. The analysis code was primarily written by NAM with input from AC, KK, and CW. The paper was primarly written by NAM with inputs from all authors.}   \vskip 5pt

Presented in Chapter~\ref{cha:lensing}.\vskip 10pt
 
 
% \item[\cite{Coogan:2022cky}] 
%A. Coogan, {N. Anau Montel}, K. Karchev,  M. W. Grootes, F. Nattino, C. Weniger, \textit{The effect of the perturber population on subhalo measurements in strong gravitational lenses}, \href{https://doi.org/10.1093/mnras/stac3215}{Mon.Not.Roy.Astron.Soc. 518 (2023) 2746}, \href{https://arxiv.org/abs/2205.09126}{\ttfamily arXiv:2205.09126}.  \vskip 5pt
% 
%{AC, NAM, KK, CW participated in the planning of the project. The analysis code was primarily written by AC with input from NAM, KK, and CW. The paper was primarly written by AC with inputs from all authors.}   \vskip 5pt
%
%
%Presented in \autoref{cha:lensing}.\vskip 10pt


\item[\cite{AnauMontel:2023stj}] 
{N. Anau Montel}, J. Alvey, C. Weniger, \textit{Scalable inference with Autoregressive Neural Ratio Estimation}, \href{https://doi.org/10.1093/mnras/stae1130}{Mon.Not.Roy.Astron.Soc. 527 (2024) 66},  \href{https://arxiv.org/abs/2308.08597}{\ttfamily arXiv:2308.08597}.  \vskip 5pt

{All authors participated in the planning of the project. The analysis code was primarily written by NAM, except for the stellar stream example contributed by JA. The paper was written by NAM and JA, with input from CW.}   \vskip 5pt

Presented in Chapter~\ref{cha:anre}. \vskip 10pt


\item[\cite{List:2023aa}] 
F. List, {N. Anau Montel}, C. Weniger, \textit{Bayesian Simulation-based Inference for Cosmological Initial Conditions}, Machine Learning and the Physical Sciences Workshop at the 37th Conference on Neural Information Processing Systems (NeurIPS 2023) \href{https://ml4physicalsciences.github.io/2023/files/NeurIPS_ML4PS_2023_218.pdf}{[Paper]} \href{https://nips.cc/media/PosterPDFs/NeurIPS%202023/76248.png}{[Poster]} ,  \href{https://arxiv.org/abs/2310.19910}{\ttfamily arXiv:2310.19910}.  \vskip 5pt

{All authors participated in the planning of the project. FL provided the simulations, NAM and FL wrote the analysis code with input from CW. The paper was written by both NAM and FL, with input from CW.}   \vskip 5pt

Presented in Chapter~\ref{cha:cosmo}. \vskip 10pt


\item[\cite{AnauMontel:2022ppb}] 
{N. Anau Montel}, C. Weniger, \textit{Detection is truncation: studying source populations with truncated marginal neural ratio estimation}, Machine Learning and the Physical Sciences Workshop at the 36th Conference on Neural Information Processing Systems (NeurIPS 2022) \href{https://ml4physicalsciences.github.io/2022/files/NeurIPS_ML4PS_2022_50.pdf}{[Paper]} \href{https://neurips.cc/media/PosterPDFs/NeurIPS%202022/57017.png}{[Poster]} , \href{https://arxiv.org/abs/2211.04291}{\ttfamily arXiv:2211.04291}.  \vskip 5pt

{Both authors participated in the planning of the project. The analysis code was written by NAM with input from CW. The paper was primarly written by NAM with input from CW.}   \vskip 5pt

Presented in Chapter~\ref{cha:detection}. \vskip 10pt

}

\end{itemize}

% Others ============================================================

{\scshape Other publications by the author:} \vskip 20pt

\begin{itemize}[]

{\fontfamily{cmr}\selectfont

 \item[\cite{Coogan:2022cky}] 
A. Coogan, {N. Anau Montel}, K. Karchev,  M. W. Grootes, F. Nattino, C. Weniger, \textit{The effect of the perturber population on subhalo measurements in strong gravitational lenses}, \href{https://doi.org/10.1093/mnras/stac3215}{Mon.Not.Roy.Astron.Soc. 518 (2023) 2746}, \href{https://arxiv.org/abs/2205.09126}{\ttfamily arXiv:2205.09126}.  \vskip 5pt

\item[\cite{Karchev:2022aa}] 
K. Karchev, {N. Anau Montel}, A. Coogan, C. Weniger, \textit{Strong-Lensing Source Reconstruction with Denoising Diffusion Restoration Models}, Machine Learning and the Physical Sciences Workshop at the 36th Conference on Neural Information Processing Systems (NeurIPS 2022) \href{https://ml4physicalsciences.github.io/2022/files/NeurIPS_ML4PS_2022_169.pdf}{[Paper]} \href{https://neurips.cc/media/PosterPDFs/NeurIPS%202022/56913.png}{[Poster]}, \href{https://arxiv.org/abs/2211.04365}{\ttfamily arXiv:2211.04365}. \vskip 5pt

\item[\cite{Correa:2022aa}] 
C. Correa, M. Schaller, S. Ploeckinger, {N. Anau Montel}, C. Weniger, S. Ando, \textit{TangoSIDM: Tantalizing models of Self-Interacting Dark Matter}, \href{https://doi.org/10.1093/mnras/stac2830}{Mon.Not.Roy.Astron.Soc. 517 (2022) 3045}, \href{https://arxiv.org/abs/2206.11298}{\ttfamily arXiv:2206.11298}. \vskip 5pt

%\item[\cite{Correa:2022aa}] 
%{N. Anau Montel}, C. Correa, \textit{TangoSIDM: the effect of dark matter self-interactions on satellite galaxies.},  \vskip 5pt


}

\end{itemize}


\newpage

